Multivariate Analysis

### 2021-02-04 multivariate.Rmd
more modeling

# final cleanup and factoring  
aoc_z$Species <- as.factor(paste(aoc_z$genus,aoc_z$species)) #must make value 'Species" (uppercase)
aoc_z$Group <- as.factor(aoc_z$organism.group) #must make value "Group"
aoc_z$Group <- fct_explicit_na(aoc_z$Group) #makes sure that species get counted even if they're missing a group
```
```{r}
# subset data to selected variables
multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    ID,
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect_f, #yes no
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm,#max ingestible size
                                   acute.chronic_f,
                                   dose.mg.L.master.AF.noec,
                                   dose.particles.mL.master.AF.noec) %>%  
  filter(!size_f == "Not Reported") %>%   #take out not reported 
  mutate_if(is.character, as.factor)

# #recode variables
# multiVar <- multiVar %>% mutate(effect_10 = case_when(
#     effect_f == "Y" ~ 1,
#     effect_f == "N" ~ 0
#   ))# 
```

@@

## Continuous Predictors
Repeat with continous variables whenever possible, and max of 8 predictors.
```{r}
skim(multiVar)
```
```{r}
acute <- multiVar %>% filter(acute.chronic_f == "Chronic") %>% 
  filter(!environment == "Terrestrial") %>% 
  filter(bio.org == "organism") %>% 
  filter(lvl1_f == "Fitness") %>% 
  filter(!exposure.route == "food") %>% 
  select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d,bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master)) %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(acute)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
plot(myrf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(myrf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- myrf$importance
View(importance)
# displays the data plotted in the plot above
```

>>>>

# final cleanup and factoring  
aoc_z$Species <- as.factor(paste(aoc_z$genus,aoc_z$species)) #must make value 'Species" (uppercase)
aoc_z$Group <- as.factor(aoc_z$organism.group) #must make value "Group"
aoc_z$Group <- fct_explicit_na(aoc_z$Group) #makes sure that species get counted even if they're missing a group
```
```{r}
# subset data to selected variables
multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    ID,
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect_f, #yes no
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm,#max ingestible size
                                   acute.chronic_f,
                                   dose.mg.L.master.AF.noec,
                                   dose.particles.mL.master.AF.noec,
                                   effect.score) %>%  #1 = minor, 2 = photosynthesis, feeding, 3 = growth, chlorophyll content, 4 = reproduction, 5  = population growth, 6 = survival
  filter(!size_f == "Not Reported") %>%   #take out not reported 
  mutate_if(is.character, as.factor) %>% mutate(effect_10 = case_when(
     effect_f == "Yes" ~ 1,
     effect_f == "No" ~ 0))
```

@@

## Continuous Predictors
Repeat with continous variables whenever possible, and max of 8 predictors.
```{r}
skim(multiVar)
```
```{r}
acute <- multiVar %>% 
  filter(acute.chronic_f == "Acute") %>% 
  filter(!environment == "Terrestrial") %>% 
  filter(bio.org == "organism") %>% 
  filter(lvl1_f == "Fitness") %>% 
  filter(!exposure.route == "food") %>% 
  dplyr::select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d, bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master, effect_10, ID)) %>%
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(acute)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
acute_split <- acute %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
acute_train <- training(acute_split)
acute_test <- testing(acute_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Random forest -- 
set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
acuterf <- randomForest(y = acute_train$effect_f, # dependent variable
  x = acute_train %>%
    dplyr::select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

acuterf # examine the results.
```

```{r}
plot(acuterf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(acuterf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- acuterf$importance
View(importance)
# displays the data plotted in the plot above
```

###Validation

```{r fit results}
fitted <- predict(acuterf, acute_test[,-6], OOB = TRUE, type ="response")
misClasificError <- mean(fitted != acute_test$effect_f)
print(paste('Training Accuracy', 1 - misClasificError))
```

Alternative ROC Curve
```{r}
require(pROC)
predicted <- predict(acuterf, acute_test %>%  dplyr::select(-effect_f),
                       OOB=TRUE, type= "response")
#Calculate ROC curve
rocCurve.tree <- roc(as.numeric(acute_test$effect_f),as.numeric(predicted))

##gplot
# rocks <- roc()

#plot the ROC curve
plot(rocCurve.tree,col=c(4))
```
##### Chronic
```{r}
chronic <- multiVar %>% 
  filter(acute.chronic_f == "Chronic") %>% 
  filter(!environment == "Terrestrial") %>% 
  filter(bio.org == "organism") %>% 
  filter(lvl1_f == "Fitness") %>% 
  filter(!exposure.route == "food") %>% 
  dplyr::select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d, bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master, effect_10, ID)) %>%
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(chronic)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
chronic_split <- chronic %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
chronic_train <- training(chronic_split)
chronic_test <- testing(chronic_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Random forest -- 
set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
chronicrf <- randomForest(y = chronic_train$effect_f, # dependent variable
  x = chronic_train %>%
    dplyr::select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

chronicrf # examine the results.
```

```{r}
plot(chronicrf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(chronicrf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- chronicrf$importance
View(importance)
# displays the data plotted in the plot above
```

```{r fit results}
fitted <- predict(chronicrf, chronic_test[,-6], OOB = TRUE, type ="response")
misClasificError <- mean(fitted != chronic_test$effect_f)
print(paste('Training Accuracy', 1 - misClasificError))
```

#### All Data
```{r}
all <- multiVar %>% 
  #filter(acute.chronic_f == "Chronic") %>% 
  filter(!environment == "Terrestrial") %>% 
  filter(bio.org == "organism") %>% 
  filter(lvl1_f == "Fitness") %>% 
  filter(!exposure.route == "food") %>% 
  dplyr::select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d, bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master, effect_10, ID)) %>%
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(all)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
all_split <- all %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
all_train <- training(all_split)
all_test <- testing(all_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Random forest -- 
set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
allrf <- randomForest(y = all_train$effect_f, # dependent variable
  x = all_train %>%
    dplyr::select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

allrf # examine the results.
```

```{r}
plot(allrf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(allrf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- allrf$importance
View(importance)
# displays the data plotted in the plot above
```

```{r fit results}
fitted <- predict(allrf, all_test[,-6], OOB = TRUE, type ="response")
misClasificError <- mean(fitted != all_test$effect_f)
print(paste('Training Accuracy', 1 - misClasificError))
```

#### NO FILTERS
```{r}
nofilter <- multiVar %>% 
  #filter(acute.chronic_f == "Chronic") %>% 
  ##filter(!environment == "Terrestrial") %>% 
  #filter(bio.org == "organism") %>% 
  #filter(lvl1_f == "Fitness") %>% 
  #filter(!exposure.route == "food") %>% 
  dplyr::select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d, bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master, effect_10, ID)) %>%
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(nofilter)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
nofilter_split <- nofilter %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" cnofilter ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
nofilter_train <- training(nofilter_split)
nofilter_test <- testing(nofilter_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Random forest -- 
set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
nofilterrf <- randomForest(y = nofilter_train$effect_f, # dependent variable
  x = nofilter_train %>%
    dplyr::select(-effect_f), # selecting nofilter predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

nofilterrf # examine the results.
```

```{r}
plot(nofilterrf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(nofilterrf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- nofilterrf$importance
View(importance)
# displays the data plotted in the plot above
```

```{r fit results}
fitted <- predict(nofilterrf, nofilter_test[,-6], OOB = TRUE, type ="response")
misClasificError <- mean(fitted != nofilter_test$effect_f)
print(paste('Training Accuracy', 1 - misClasificError))
```

#### ROC Curve
yet anothyer try
```{r}
###CHRONIC
chronicpredictions <- as.data.frame(predict(chronicrf, chronic_test[,-6], type = "prob"))
# predict class and then attach test class
chronicpredictions$predict <- names(chronicpredictions)[1:2][apply(chronicpredictions[,1:2], 1, which.max)]
chronicpredictions$observed <- chronic_test$effect_f

####ACUTE
predictions <- as.data.frame(predict(acuterf, acute_test[,-6], type = "prob"))
# predict class and then attach test class
predictions$predict <- names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
predictions$observed <- acute_test$effect_f
head(predictions)

#### all
allpredictions <- as.data.frame(predict(allrf, all_test[,-6], type = "prob"))
# predict class and then attach test class
allpredictions$predict <- names(allpredictions)[1:2][apply(allpredictions[,1:2], 1, which.max)]
allpredictions$observed <- all_test$effect_f
head(allpredictions)

###nofilter
nofilterpredictions <- as.data.frame(predict(nofilterrf, nofilter_test[,-6], type = "prob"))
# predict class and then attach test class
nofilterpredictions$predict <- names(nofilterpredictions)[1:2][apply(nofilterpredictions[,1:2], 1, which.max)]
nofilterpredictions$observed <- nofilter_test$effect_f
```
```{r}
# 1 ROC curve, yes vs no for acute
roc.acute <- roc(ifelse(predictions$observed=="Yes", "Yes", "No"), as.numeric(predictions$Yes))

#chronic
roc.chronic <- roc(ifelse(chronicpredictions$observed=="Yes", "Yes", "No"), as.numeric(chronicpredictions$Yes))
#all
roc.all <- roc(ifelse(allpredictions$observed=="Yes", "Yes", "No"), as.numeric(allpredictions$Yes))

#nofilter
roc.nofilter <- roc(ifelse(nofilterpredictions$observed=="Yes", "Yes", "No"), as.numeric(nofilterpredictions$Yes))

all <- ggroc(roc.all) + labs(title = "Chronic and Acute",
                             subtitle = "Accuracy = 81.7%, n = 1510") +
  theme_minimal()
acute <- ggroc(roc.acute, col = "red") + labs(title = "Acute",
                                              subtitle = "Accuracy = 84.8%, n = 883") + theme_minimal()
chronic <- ggroc(roc.chronic, col = "blue") + labs(title = "Chronic",
                                                   subtitle = "Accuracy = 67.3%, n = 452") + theme_minimal()

nofilter<- ggroc(roc.nofilter, col = "green") + labs(title = "Entire Dataset",
                                                   subtitle = "Accuracy = 79.9%, n = 4615") + theme_minimal()
require(gridExtra)
grid.arrange(all, acute, chronic, nofilter,
             ncol = 4)
```

```{r}
#plot reciever operating curve
perf <- performance(preds,"tpr","fpr")
```

```{r}
# compute in-sample results
caret::confusionMatrix(fitted,as.factor(acute_test$effect_f))
```

```{r}
# Validation set assessment #1: looking at confusion matrix
prediction_for_table <- predict(acuterf, acute_test[,-6])
table(observed = acute_test[,6], predicted = prediction_for_table)
```

```{r}
# Validation set assessment #2: ROC curves and AUC
# Needs to import ROCR package for ROC curve plotting:
library(ROCR)
# Calculate the probability of new observations belonging to each class
# prediction_for_roc_curve will be a matrix with dimensions data_set_size x number_of_classes
prediction_for_roc_curve <- predict(acuterf ,acute_test[,-6],type="response")
# Use pretty colours:
pretty_colours <- c("#F8766D","#00BA38","#619CFF")
# Specify the different classes 
classes <- levels(acute_test$organism.group)
# For each class
for (i in 1:13)
{
 # Define which observations belong to class[i]
 true_values <- ifelse(acute_test[,6]==classes[i],"Yes","No")
 # Assess the performance of classifier for class[i]
 pred <- prediction(prediction_for_roc_curve[,i],true_values)
 perf <- performance(pred, "tpr", "fpr")
 if (i==1)
 {
     plot(perf,main="ROC Curve",col=pretty_colours[i]) 
 }
 else
 {
     plot(perf,main="ROC Curve",col=pretty_colours[i],add=TRUE) 
 }
 # Calculate the AUC and print it to screen
 auc.perf <- performance(pred, measure = "auc")
 print(auc.perf@y.values)
}
```
